---
output: github_document
---

# Class 9
# Subsetting, working with data frames and `dplyr` verbs

## Subsetting

Data frame can be subsetted with the use of coordinates (indexes) enclosed within the square brackets. In case of data frames there are always 2 coordinates: `[row_number,column_number]`.  
Note that vectors have only a single  coordinate, the position, so they are one-dimensional objects. Data frames have two dimensions: rows and columns. 

#### Exercise 1
Return the value from 49^th^ row and 4^th^ column of `my_data` from Class 8 (use your script/history to get this variable)

Expected result:  
```{r echo = FALSE, comment= ""}
CO2[49,4]
```

--- 

You can call multiple values at the same time. To define a range of coordinates, use colon to separate range limits: `[row_number_1:row_number_2, column_number]`.

#### Exercise 2
Return first 5 rows of the last 3 columns of `my_data`. To check the number of rows and columns use `str()`, which will give you also extra info about the data frame or `dim()` which will return just the number of rows and columns

Expected result:
```{r echo = FALSE, comment= ""}
CO2[1:5,3:5]
```

---

If you want to call for all rows or columns it is enough to put nothing instead of the respective coordinate, e. g. `[,column_number]`

#### Exercise 3
Now, save the 5^th^ column of `my_data` as a variable called `my_column_5`. Call call the variable, so its content is displayed in your console.

Expected result:
```{r echo = FALSE, comment= ""}
CO2[,5]
```

---

Note that the result is no longer a table. As we asked for just one column, a series of numbers (vector) was returned. Let's recall vector subsetting.

#### Exercise 4
Return 3^rd^, 4^th^ and 5^th^ value from `my_column_5`.

Expected result:
```{r echo = FALSE, comment= ""}
CO2[3:5,5]
```

---

> #### Important 
Pulling one row from the data frame will not result in a vector. It's because a row can contain elements of different types, which is not allowed for vectors.

Table can be also subsetted with the use of column (or row, if they exist) names.

#### Exercise 5
Return the same column as in Ex. 4, but use column name instead of its number. Don't forget about quotation marks.

Another way of obtain an entire column is to use dollar sign between data frame and column names `dataframe_name$column_name`. Such expression is automatically treated as a vector, so it can be directly subsetted to get a particular row: `dataframe_name$column_name[row_number]`.

#### Exercise 6
Return values of rows 15^th^ to 30^th^ of `conc` column from `my_data`. Use `$` sign.

Expected result:
```{r echo = FALSE, comment= ""}
CO2$conc[15:30]
```

---

If desired rows (or columns) do not follow each other and the range option cannot be used, vector of coordinates should be provided.

#### Exercise 7
Create a vector of integers with values 3, 5, 7, 9 and 12 and save it to a variable. Call it.

Expected result:
```{r echo = FALSE, comment= ""}
c(3,5,7,9,12)
```

---

#### Exercise 8
Return rows of `my_data` `uptake` column corresponding to the numbers in the vector created in the previous exercise.

Expected result:
```{r echo = FALSE, comment= ""}
CO2$uptake[c(3,5,7,9,12)]
```

---

You can also select everything except specified columns (or rows), i. e., drop rows or columns. To do this, put minus (-) before an index or vector of indexes.

#### Exercise 9

Return all columns of `my_data` except 5^th^.

First 10 output lines are shown below: 
```{r echo = FALSE, comment= ""}
head(CO2[,-5],10)
```

---

## Simple operations on data frames

### Replacement

Assign a particular value to a specific position in your your data frame with the use of the assignment arrow `<-`. It works just as with variable assignment: `data_frame[row_number,column_number] <- new_value`

#### Exercise 10
Replace 5^th^ value in the column `Type` of `my_data` with the label 'Unknown'. Check whether this was indeed replaced.

---

### Mathematical operations

You can use classical mathematical operators: `+`, `-`, `*` and `/`. Remember, however, that mathematical operations make sense only for `numeric` data types.

#### Exercise 11
The `conc` column gives CO2 concentration in ul/L, recalculate concentration to ml/L and modify the column. Display the modified column.

Expected result:
```{r echo = FALSE, comment= ""}
CO2$conc <- CO2$conc/1000
CO2$conc
```

---

Also, you can use simple summary functions from the previous class (see Class 8).

#### Exercise 12
Calculate the mean value of the `uptake` column.

Expected result:
```{r echo = FALSE, comment= ""}
mean(CO2[,5])
```

---

### Simple summaries
- `nrow()` - number of rows
- `ncol()` - number of columns
- `summary()` - descriptive statistics for each column

#### Exercise 13
Return the total number of cells within in `my_data`.

Expected result:
```{r echo = FALSE, comment= ""}
nrow(CO2)*ncol(CO2)
```

#### Exercise 14
Display summary of `my_data`. Note the the type of summary depends on the type of data in a column.

Expected result:
```{r echo = FALSE, comment= ""}
summary(CO2)
```


### Deleting rows containing missing data

Missing data, as explained in the previous class, are represented as `NA` (Not Available) in R. Many functions will raise an error every time it encounters `NA` as their arguments or their parts. Many functions, on the other hand, allow you to automatically discard `NA`. The default behaviour of not ignoring `NA` is by design, as `NA` actually carry important information about your data.

#### Exercise 15
Choose one cell of `my_data` and replace its value with `NA`. Do not add quotation marks as `NA` is an internal R symbol (just as $\pi$). Print the entire row.

---

Rows with missing data can be removed with `na.omit()` function. To save changes, the result of `na.omit()` function must be assigned to variable. **In practice, as explained above, deleting missing data must be well justified**.

#### Exercise 16
Remove the row with missing data created in the previous exercise. Replace `my_data` with such modified table. Remember that this action cannot be  undone.

### Adding new column or row

Adding **new column** is relatively **simple**. All you need to have is a vector. However, remember three things:

- length of vector must equal the number of rows of a data frame
- order of values within a vector corresponds to the row numbers
- name of the vector will become the name of the added column

We are going to add a column with IDs of observations. Note that column with IDs is often necessary in statistical analysis. It is also inherent to the data in [long format](https://www.theanalysisfactor.com/wide-and-long-data/) which is strongly advised (see class 1 and next class).

#### Exercise 17
Create a vector with ID numbers starting from 100. Use one of functions introduced above to determine what should be the length of the vector. Name the vector `ID`.

Expected result:
```{r echo = FALSE, comment= ""}
c(100:(98+nrow(CO2)))
```

---

You can combine data frames with the use of `cbind()` function in the following manner: `cbind(data_frame1, data_frame2)`. Note that a vector can be considered a data frame with a single column. 

#### Exercise 18

Place created IDs at the beginning (as first column) of `my_dat`. Overwrite `my_data` variable.

---

Adding **new row** is more **complicated** as you cannot create a vector with different types of data. Firstly you need to create a new data frame (similar to the old one) consisting of only new row (or rows). To do this use `data.frame()` function in the following manner: `data.frame(columnName1 = value1, columnName2 = value2,...)`.

#### Exercise 19

Create data frame with a single row and the following values: `1,'Kr1','Krakow',unknown',1000,20`. Column names should correspond to these of `my_data`. Name the newly created data frame `Added_rows`.

Expected result:
```{r echo = FALSE, comment= ""}
data.frame("ID" = 1,"Plant" = "Kr1","Type" = "Krakow","Treatment" = "unknown","conc" = 1000,"uptake" = 20)
```

---

To create a data frame with multiple rows, replace the value for each column with a vector.

To combine data frames by rows use `rbind()` function in the following manner: `rbind(data_frame1, data_frame2)`

#### Exercise 20

Add the row created in Ex. 19 at the end of `my_data`. Overwrite `my_data` variable.

---

### Saving data frame

To save your data frame into a text file use `write.table()` function. 

#### Exercise 21
Save the modified `my_data` int a `.csv` file. Consult `write.table()` manual to set the arguments. Include your surname in the file name.

---
## Logical expressions and subsetting with logical vectors

## `dplyr` and `tidyverse`

### Selecting variables with `select()`

### Filtering rows with `filter()`

### Sorting with `arrange()`

### Creating new variables with `mutate()`
