---
title: "Class 12"
author: "Szymon Drobniak"
date: "`r Sys.Date()`"
output:
  md_document:
    toc: true
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Class 12

**In the below instructions...**

**EXERCISE 1:** are bits of code to execute/practice pieces to do, often with only hints on how to perform them.

**Output** indicates the typical output you should expect from a given instruction.

## Data

We will work on a simple dataset on cholesterol levels from patients. The data presents cholesterol concentrations in plasma in patients before diet, and after 4 & 8 weeks of diet containing one of two types of margarine. The age group of patients is also indicated. Data can be downloaded here: https://raw.githubusercontent.com/wbabik/Practical_computing/teaching/Class_10/data/Cholesterol_Age_R.csv

```{r echo = T}
data_chol <- read.table('https://raw.githubusercontent.com/wbabik/Practical_computing/teaching/Class_10/data/Cholesterol_Age_R.csv',
                       sep = ';', header = T,
                       stringsAsFactors = T)

head(data_chol)
summary(data_chol)
```



## `ggplot2` scatterplot

```{r echo = T, message = F}
library(ggplot2)
```

**EXERCISE 1:** Make a scatterplot similar to one of the previous excersises, mapping the concentrations of cholesterol on the x and y axes. Use blue squares as points. You may want to use the `cex` option to increase the default symbol size (`cex` defines a multiplicative coefficient, that increases or decreases plot elements given number of times).

**Output**

```{r}
plot1 <- ggplot(data = data_chol, mapping = aes(x = Before, y = After8weeks)) +
  geom_point(color = 'blue', shape = 15, cex = 3)
plot1
```

**EXERCISE 2:** Let's improve the plot by removing the annoying gray background. add the `theme_...()` call to the plot (you can review different predefined versions of it here  <https://ggplot2.tidyverse.org/reference/ggtheme.html>) to produce a cleaner graph.
**Output**

```{r}
plot1 <- ggplot(data = data_chol, mapping = aes(x = Before, y = After8weeks)) +
  geom_point(color = 'blue', shape = 15, cex = 3) + theme_bw()
plot1
```

**EXERCISE 3:** An even more aesthetically-pleasing plot can be produced using the "classic" theme. Try also, by addind the `theme()` definition to the plot, to modify the `text` element using the following formatting: `element_text(size = 20)` - which should increase the default font sizing).
**Output**

```{r}
plot1 <- ggplot(data = data_chol, mapping = aes(x = Before, y = After8weeks)) +
  geom_point(color = 'blue', shape = 15, cex = 3) + theme_classic() + theme(text = element_text(size = 20))

### LUB

plot1 <- ggplot(data = data_chol, mapping = aes(x = Before, y = After8weeks)) +
  geom_point(color = 'blue', shape = 15, cex = 3) + theme_classic(base_size = 20)

### rekomendowana taktyka to modyfikacja warstwy theme()

plot1
```

**EXERCISE 4:** Add the `geom_smooth` aesthetic to the plot, selecting the `lm` method as its option. Do you know what does `lm` indicate?
**Output**

```{r}
plot1 <- ggplot(data = data_chol, mapping = aes(x = Before, y = After8weeks)) +
  geom_point(color = 'blue', shape = 15, cex = 3) + theme_classic() + geom_smooth(method = 'lm') +
  theme(text = element_text(size = 20))
plot1
```
**EXERCISE 5:** We can modify the above call to change the appearance of the regression line.
**Output**

```{r echo = T}
plot1 <- ggplot(data = data_chol, mapping = aes(x = Before, y = After8weeks)) +
  geom_point(color = 'blue', shape = 15, cex = 3) + theme_classic() + geom_smooth() +
  theme(text = element_text(size = 20))
plot1
```

## Annotating the plot, other dimensions of data

**EXERCISE 6:** Try to map the age groups (`AgeGroup`) to the graph points' colours. Using the `alpha` option (taking values 0 - 1 - which indicates the transparency of the regression error band) I decreased the cluttered appearance of the plot, making the error bands a bit more subtle.
**Output**

```{r}
plot1 <- ggplot(data = data_chol, mapping = aes(x = Before, y = After8weeks, color = AgeGroup)) +
  geom_point(shape = 15, cex = 3) + theme_classic() + geom_smooth(method = 'lm', alpha = 0.25) +
  theme(text = element_text(size = 20))
plot1
```

**EXERCISE 7:** let's add the `labs()` layer with a bit more readable axes names.
**Output**

```{r}
plot1 <- ggplot(data = data_chol, mapping = aes(x = Before, y = After8weeks, color = AgeGroup)) +
  geom_point(shape = 15, cex = 3) + theme_classic() + geom_smooth(method = 'lm', alpha = 0.25) +
  labs(x = "Concentration before experiment", y = "Concentration after 8 weeks", color = "Age") +
  theme(text = element_text(size = 20))
plot1
```
**EXERCISE 8:** An alternative way of coding age gropups - instead of mapping it to colours - may be splitting the groups by so called *facets*, which presents subsets of data on separate subplots using a common scale. To achieve this you should use the `facet_wrap()` function, which takes a formula of the form `~ A`, where A indicates a variable from the dataset that defines the split of the graph area into subplots. (An analogous function `facet_grid()` handles well two-sided formulas `A ~ B` that define a grid of plots). Try to recreate the below plot - it may look better with the `theme_bw()` style, instead of the "classic' one.
**Output**

```{r}
plot1 <- ggplot(data = data_chol, mapping = aes(x = Before, y = After8weeks, color = AgeGroup)) +
  facet_wrap( ~ AgeGroup) +
  geom_point(shape = 15, cex = 3) + theme_bw() + geom_smooth(method = 'lm') +
  labs(x = "Concentration before experiment", y = "Concentration after 8 weeks", color = "Age") +
  theme(text = element_text(size = 20))
plot1
```

## `ggplot2` histogram

**EXERCISE 9:** Using the `geom_hist()` geometry create a histogram of the `After8weeks` variable.
**Output**

```{r}
plot1 <- ggplot(data = data_chol, mapping = aes(x = After8weeks)) +
  geom_histogram(fill = 'white', color = 'black', bins = 10) +
  labs(x = "Concentration after 8 weeks") + theme_classic() +
  theme(text = element_text(size = 20))
plot1
```

**EXERCISE 10:** Change the histogram so that it displays relative frequencies of data in each bin, and not absolute counts. Inspiration on how to do this can be found here: <https://homepage.divms.uiowa.edu/~luke/classes/STAT4580/histdens.html> - there are at least two ways of achieving this goal!

**Output**

```{r}
### this is one of the methods (inside the geom_hist())
plot1 <- ggplot(data = data_chol, mapping = aes(x = After8weeks)) +
  geom_histogram(fill = 'white', color = 'black', bins = 10, aes(y = ..count../sum(..count..))) +
  labs(x = "Concentration after 8 weeks") +theme_classic() +
  theme(text = element_text(size = 20))
plot1
```

**EXERCISE 11:** Modify the histogram to add a kernel density estimator to it (it is an analogue of the `density()` function we have used earlier).

**Output**

```{r}
### This is the second way of turning counts to frequencies in a ggplot2 histogram
plot1 <- ggplot(data = data_chol, mapping = aes(x = Before, y = ..density..)) +
  geom_histogram(fill = 'white', color = 'black', bins = 10) +
  geom_density(color = 'red', fill = 'red', alpha = 0.15) +
  labs(x = "Concentration after 8 weeks") + theme_classic() +
  theme(text = element_text(size = 20))
plot1
```

## Grid of plots

If our goal is to place multiple unrelated plots on one figure - we may as well use the `ggpubr` package and its `ggarrange()` function.

```{r echo = T}
# install.packages('ggpubr') # if needed
library(ggpubr)
```

**EXERCISE 12:** First - produce 4 plots and save them to 4 differently named objects. These can be some of the plots we have generated so far:

```{r echo = T, message = F}
plot1 <- ggplot(data = data_chol, mapping = aes(x = Before, y = After8weeks)) +
  geom_point(color = 'blue', shape = 15, cex = 3) + theme_classic() +
  theme(text = element_text(size = 12))

plot2 <- ggplot(data = data_chol, mapping = aes(x = Before, y = ..density..)) +
  geom_histogram(fill = 'white', color = 'black', bins = 10) +
  geom_density(color = 'red', fill = 'red', alpha = 0.15) +
  theme_classic() +
  theme(text = element_text(size = 12))

plot3 <- ggplot(data = data_chol, mapping = aes(x = Before, y = After4weeks, color = AgeGroup)) +
  geom_point(shape = 15, cex = 3) + theme_classic() + geom_smooth(method = 'lm', alpha = 0.25) +
  theme(text = element_text(size = 12))

plot4 <- ggplot(data = data_chol, mapping = aes(x = After4weeks, y = After8weeks, color = AgeGroup)) +
  geom_point(shape = 15, cex = 3) + theme_classic() + geom_smooth(method = 'lm', alpha = 0.25) +
  theme(text = element_text(size = 12))

```

Now - produce a grid plot from the four subplots using the `ggarrange()` function. Label the subplots as `A`, `B`, `C`, `D`.

```{r}
gridplot <- ggarrange(plot1, plot2, plot3, plot4,
                      labels = c('A', 'B', 'C', 'D'), legend = 'bottom', common.legend = T)
gridplot
```


## Boxplot

**EXERCISE 13:** `geom_boxplot()` can be used to visualise categorical data. In base R this is achieved by using the `boxplot()` function:

```{r echo = T}

boxplot(Before ~ AgeGroup, data = data_chol,
        xlab = "Age group", ylab = "Cholesterol conc. before the experiment")

```

Try to produce such plot, showing the cholesterol concentrations before the diet, categorised by age groups, in `ggplot2`. Use `?geom_boxplot` and if needed the book <https://ggplot2-book.org/collective-geoms.html> to find out how to achieve this. On such a boxplot - what is the meaning of: the boundaries of each box, the ends of the whiskers and the additional points added to the plot?

**Output**

```{r}
plot2 <- ggplot(data = data_chol, mapping = aes(x = AgeGroup, y = Before)) +
  geom_boxplot() +
  labs(x = "Age Group", y = "Cholesterol before experiment") +
  theme_classic() + theme(text = element_text(size = 25))
plot2
```

**EXERCISE 14:** a boxplot may be more informative if we add raw data to it. It can be done in many ways - e.g., to achieve an effect similar to this one: <https://bit.ly/31estrN>. Try to produce a similar plot using additional data (file `Diet_R.csv`, which presents weight loss of patients on three different diets). Before using the data clean it from all missing values (`na.omit()`). Data can be found here: https://raw.githubusercontent.com/wbabik/Practical_computing/teaching/Class_10/data/Diet_R.csv

```{r echo = T}
mydata <- na.omit(read.table("https://raw.githubusercontent.com/wbabik/Practical_computing/teaching/Class_10/data/Diet_R.csv", sep = ",", header = T))
```

To achieve the below effect remember to include the following components: use `geom_boxplot()` and `geom_point()` to map data to visuals; map genders to box colours; for both points and boxes you can set `alpha` option to something <1 to ensure that they are a bit teansparent and hence less tiring to our eyes.

**Output**

```{r}
plot3 <- ggplot(data = mydata, mapping = aes(x = as.factor(Diet),
                                            color = as.factor(gender),
                                            y = weight6weeks,
                                            fill = as.factor(gender))) +
  geom_boxplot(alpha = 0.3, outlier.shape = NA) + 
  theme_classic() +
  geom_point(color = "black",
             position = position_jitterdodge(jitter.width = 0.1),
             size = 3, alpha = 0.5, stroke = 0) +
  theme(text = element_text(size = 25)) +
  labs(x = "Diet type", y = "Weight after 6 weeks",
       color = "Gender", fill = "Gender")

plot3
```

## Customizing plot elements - scales, axes, legends

In our previous example we used in all cases default colours supplied by `ggplot`. Let's explore the ways we can modify legends. First - produce the below plot based on the `Diet_R.csv' data, it shows the relationship between body height and post-diet weight in three different diet groups.

```{r echo = T, message = F}
myplot <- ggplot(data = mydata, mapping = aes(x = Height, y = weight6weeks, colour = as.factor(Diet))) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_classic() +
  theme(text = element_text(size = 20)) +
  labs(x = "Height", y = "Weight after 6 weeks", colour = "Diet type", title = "Height vs. weight")
myplot
```

**EXERCISE 15:** add anothe layer to the plot - using the `scale_colour_manual()` component change colours used on the plot to purple, organe and dark grey for diets 1, 2 and 3. In order to achieve this you should assign these colours to `values` of the scale.

**Output**

```{r message = F}
myplot <- ggplot(data = mydata, mapping = aes(x = Height, y = weight6weeks, colour = as.factor(Diet))) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_classic() +
  theme(text = element_text(size = 20)) +
  labs(x = "Height", y = "Weight after 6 weeks", colour = "Diet type", title = "Height vs. weight") +
  scale_colour_manual(values = c('purple', 'orange', 'darkgrey'))
myplot
```

**EXERCISE 16:** to give some more space for the width of the plot move the legend (by modifying the `theme()` function call) to the bottom of the plot.

**Output**

```{r message = F}
myplot <- ggplot(data = mydata, mapping = aes(x = Height, y = weight6weeks, colour = as.factor(Diet))) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_classic() +
  theme(text = element_text(size = 20), legend.position = "bottom") +
  labs(x = "Height", y = "Weight after 6 weeks", colour = "Diet type", title = "Height vs. weight") +
  scale_colour_manual(values = c('purple', 'orange', 'darkgrey'))
myplot
```

**EXERCISE 17:** by specifying `guide = "none"` in the scale layer, you can remove the legend altogether - try it.

**Output**

```{r message = F}
myplot <- ggplot(data = mydata, mapping = aes(x = Height, y = weight6weeks, colour = as.factor(Diet))) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_classic() +
  theme(text = element_text(size = 20)) +
  labs(x = "Height", y = "Weight after 6 weeks", colour = "Diet type", title = "Height vs. weight") +
  scale_colour_manual(values = c('purple', 'orange', 'darkgrey'), guide = "none")
myplot
```

**EXERCISE 18:** by using the `labels` option in the scale layer you can easily rename the categories presented in the lgend. Try renaming them to vegan (diet 1), lacto-ovo (diet 2) and vegetarian (diet 3).

**Output**

```{r message = F}
myplot <- ggplot(data = mydata, mapping = aes(x = Height, y = weight6weeks, colour = as.factor(Diet))) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_classic() +
  theme(text = element_text(size = 20), legend.position = "bottom") +
  labs(x = "Height", y = "Weight after 6 weeks", colour = "Diet type", title = "Height vs. weight") +
  scale_colour_manual(values = c('purple', 'orange', 'darkgrey'),
                      labels = c('vegan', 'lacto-ovo', 'vegetarian'))
myplot
```
**EXERCISE 19:** similarly, you can modify colour scales of any other component - e.g., in the plot...

```{r echo = T}
myplot <- ggplot(data = mydata, mapping = aes(x = as.factor(Diet),
                                            color = as.factor(gender),
                                            y = weight6weeks,
                                            fill = as.factor(gender))) +
  geom_boxplot(alpha = 0.3, outlier.shape = NA) + 
  labs(x = "Diet type", y = "Weight after diet", colour = "Gender", fill = "Gender") +
  theme_classic()
myplot
```

...we can modify scales linked to colour, fill and, for example, x axis. Use `scale_colour_viridis_d()` (a discrete version of the viridis palette), `scale_fill_viridis_d` and `scale_x_discrete` to modify the above plot and reproduce the below example. Note: if you don;t have the `viridis` library installed, execute first `install.packages('viridis')`.

```{r}
myplot <- ggplot(data = mydata, mapping = aes(x = as.factor(Diet),
                                            color = as.factor(gender),
                                            y = weight6weeks,
                                            fill = as.factor(gender))) +
  geom_boxplot(alpha = 0.3, outlier.shape = NA) + 
  labs(x = "Diet type", y = "Weight after diet", colour = "Gender", fill = "Gender") +
  scale_colour_viridis_d() +
  scale_fill_viridis_d() +
  scale_x_discrete(labels = c('vegan', 'lacto-ovo', 'vegeterian')) +
  theme_classic() +
  theme(text = element_text(size = 20))
myplot
```

**EXERCISE 20:** similarly to using the `xlim`/`ylim` options in the `plot()` function, you can modify the plotting area of a `ggplot` graph. There are two way to achieve this. The first involves adding the `xlim()` (check `?xlim` for details) or `ylim()` layers (or both). The other works by modifying the coordinate system used and adding the layer: `coord_cartesian(xlim = c(...))`. Which produces which plot below? How do they differ in handling the plotted data?

**Output**

```{r message = F}
myplot <- ggplot(data = mydata, mapping = aes(x = Height, y = weight6weeks, colour = as.factor(Diet))) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_classic() +
  theme(text = element_text(size = 20), legend.position = "bottom") +
  labs(x = "Height", y = "Weight after 6 weeks", colour = "Diet type", title = "Height vs. weight") +
  scale_colour_manual(values = c('purple', 'orange', 'darkgrey'),
                      labels = c('vegan', 'lacto-ovo', 'vegetarian')) +
  xlim(c(160, 175))
myplot

myplot <- ggplot(data = mydata, mapping = aes(x = Height, y = weight6weeks, colour = as.factor(Diet))) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_classic() +
  theme(text = element_text(size = 20), legend.position = "bottom") +
  labs(x = "Height", y = "Weight after 6 weeks", colour = "Diet type", title = "Height vs. weight") +
  scale_colour_manual(values = c('purple', 'orange', 'darkgrey'),
                      labels = c('vegan', 'lacto-ovo', 'vegetarian')) +
  coord_cartesian(xlim = c(160, 175))
myplot
```


## Saving the plots

Use the `ggsave` function - it provides some powerful plot saving routines.

**EXERCISE 21:** Save one of the last plots to a JPG file and to a PDF file. Rescale the pdf to about 80% of the original plot size. Check `?ggsave` for more information. Compare what happens when you save the plot to a PDF file directly from the plotting console.

